---
title: "Amino acids and genomic coordinates"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Amino acids and genomic coordinates}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

Resistance markers are named by residue — *pfcrt* K76T, *pfdhps* A437G, *pfkelch13* C580Y —
while callsets, scans and every plot in this package work in genomic coordinates. Going
between the two is not arithmetic you can do in your head: it depends on the transcript's
exon structure, which strand it is on, and the CDS phase.

Two functions do it, in opposite directions:

- `aa_intervals()` — a residue → the genomic interval of its codon
- `snp_aa_positions()` — a SNP → the residue it sits on, or `NA` if it is not coding

Both read coding exons from a GFF via `read_gff_cds()`.

```{r setup}
library(plasgenomicsutilsR)
```

## Getting an annotation

`read_gff_cds()` takes a path or a URL, and will read a gzipped file straight from the web,
so a released annotation can be used without keeping a copy:

```{r sources, eval = FALSE}
# VEuPathDB / PlasmoDB — the source the bundled gene datasets were built from
cds <- read_gff_cds(paste0("https://plasmodb.org/common/downloads/Current_Release/",
                           "Pfalciparum3D7/gff/data/PlasmoDB-68_Pfalciparum3D7.gff"))

# Ensembl Protists, gzipped
cds <- read_gff_cds(paste0("https://ftp.ensemblgenomes.ebi.ac.uk/pub/protists/current/",
                           "gff3/plasmodium_falciparum/",
                           "Plasmodium_falciparum.GCA000002765v3.63.gff3.gz"))
```

The examples here use a small file that ships with the package instead: the real CDS
features of six 3D7 drug-resistance genes, taken verbatim from PlasmoDB. Real coordinates,
so every number below is the published one.

```{r fixture}
cds <- read_gff_cds(system.file("extdata", "pf3d7_drug_gene_cds.gff",
                                package = "plasgenomicsutilsR"))
cds
```

## A residue to its codon

Six markers you would recognise from a resistance table:

```{r aa-to-genome}
markers <- data.frame(
  transcript_id = c("pfcrt", "pfdhps", "pfdhps", "pfdhps", "pfdhfr", "pfmdr1", "pfkelch13"),
  aa_position   = c(     76,      437,      540,      581,      108,       86,         580))

aa_intervals(markers, cds, one_based_output = TRUE)[
  , c("name", "chr", "start", "end", "codon_positions", "strand")]
```

*pfcrt* codon 76 lands at 403,624–403,626 and *pfkelch13* codon 580 at 1,725,258–1,725,260 —
the positions these mutations are reported at.

`transcript_id` took gene symbols above because `genes` defaults to `PF3D7_GENES`. A
transcript id (`"PF3D7_0709000.1"`) or a gene id (`"PF3D7_0709000"`, returning every
transcript) works the same way.

Coordinates come back **0-based half-open** by default, matching every other interval in the
package (`?"plasgenomicsutilsR-coordinates"`), which is what lets the result be used as an
interval table directly. `one_based_output = TRUE`, used above, gives the 1-based numbers you
would quote in text.

### When a codon straddles an intron

*pfcrt* has 13 coding exons, and four of its codons are split across two of them:

```{r spans-intron}
pfcrt_all <- aa_intervals(data.frame(transcript_id = "pfcrt", aa_position = 1:424), cds,
                          one_based_output = TRUE)
subset(pfcrt_all, spans_intron)[, c("name", "start", "end", "codon_positions")]
```

Codon 178 is bases 404,109 and 404,110 at the end of one exon, then 404,283 at the start of
the next. `start` and `end` span the intron with them, so that interval is 175 bp wide rather
than 3 — `spans_intron` flags it, and `codon_positions` always lists the three bases
themselves.

### On the minus strand

*pfkelch13* is transcribed right to left, so its transcript is read from the highest
coordinate down. Codon 580's first base is its *highest* genomic position:

```{r minus-strand}
snp_aa_positions(data.frame(chr = "Pf3D7_13_v3", pos = c(1725260, 1725259, 1725258)),
                 cds, keep = "hits", one_based_snps = TRUE)[
  , c("pos", "aa_position", "codon_base", "strand")]
```

## A SNP to its residue

The other direction, on positions around *pfdhps* codon 437:

```{r genome-to-aa}
snp_aa_positions(data.frame(chr = "Pf3D7_08_v3",
                            pos = c(549684, 549685, 549686, 548400, 550700)),
                 cds, one_based_snps = TRUE)[
  , c("pos", "gene_id", "aa_position", "codon_base", "coding")]
```

`aa_position` is **1-based**, counting the initiator methionine as 1 — the one deliberate
exception to this package's 0-based rule, at both ends: where you supply a residue and where
one is reported. Residue numbering is too settled to renumber (*pfcrt* K76T is codon 76
wherever it is written down, and calling it 75 would be wrong in the only sense that
matters). The genomic `start`/`end` in the same table stay 0-based half-open;
`codon_positions` is the one place a 1-based *position* is reported, since those are the
numbers people quote.

`codon_base` says which of the codon's three bases the SNP is, in transcript orientation.
`548400` is intronic and `550700` past the CDS, so both are non-coding and come back `NA`;
`keep = "hits"` drops them instead.

The two functions are exact inverses:

```{r round-trip}
iv <- aa_intervals(markers, cds, one_based_output = TRUE)
mid <- as.integer(vapply(strsplit(iv$codon_positions, ","), `[`, character(1), 2))
back <- snp_aa_positions(data.frame(chr = iv$chrom, pos = mid), cds, keep = "hits",
                         one_based_snps = TRUE)
all(back$aa_position == markers$aa_position)
```

Being inside a codon is not the same as changing the residue: that needs the alleles, which
neither function has.

## The reference residue

What the reference *does* say is which residue the codon codes for today. Hand
`snp_aa_positions()` some sequence and it adds `ref_codon` and `ref_aa`. `fasta =` takes the
same path-or-URL as `read_gff_cds()`, so point it at the released genome:

```{r ref-aa, eval = FALSE}
cds <- read_gff_cds(paste0("https://plasmodb.org/common/downloads/Current_Release/",
                           "Pfalciparum3D7/gff/data/PlasmoDB-68_Pfalciparum3D7.gff"))
genome <- paste0("https://plasmodb.org/common/downloads/Current_Release/",
                 "Pfalciparum3D7/fasta/data/PlasmoDB-68_Pfalciparum3D7_Genome.fasta")

markers <- data.frame(
  transcript_id = c("pfcrt", "pfdhps", "pfdhps", "pfdhfr", "pfmdr1", "pfkelch13"),
  aa_position   = c(     76,      437,      581,      108,       86,         580))
codons <- aa_intervals(markers, cds, one_based_output = TRUE)
middle <- as.integer(vapply(strsplit(codons$codon_positions, ","), `[`, character(1), 2))

snp_aa_positions(data.frame(chr = codons$chrom, pos = middle), cds, keep = "hits",
                 one_based_snps = TRUE, fasta = genome)[
  , c("gene_id", "aa_position", "ref_codon", "ref_aa", "strand")]
```

```
#> # A tibble: 6 × 5
#>   gene_id       aa_position ref_codon ref_aa strand
#>   <chr>               <int> <chr>     <chr>  <chr>
#> 1 PF3D7_0709000          76 AAA       K      +
#> 2 PF3D7_0810800         437 GGT       G      +
#> 3 PF3D7_0810800         581 GCG       A      +
#> 4 PF3D7_0417200         108 AGC       S      +
#> 5 PF3D7_0523000          86 AAT       N      +
#> 6 PF3D7_1343700         580 TGT       C      -
```

Read down `ref_aa`: K76, A581, S108, N86, C580 — the residue each of those markers is named
for. **Except *pfdhps* 437, which comes back `G`.** That is not an error: 3D7 itself carries the
437G allele, so a callset aligned to 3D7 shows no variant at the position everyone calls A437G.

That is worth internalising rather than filing as a curiosity. A reference is one isolate's
genome — not a consensus, and not the ancestral or wild-type sequence — and across *Plasmodium*
species the reference is sometimes the non-wild-type allele. "REF" means "what this isolate has",
never "what came first", so a marker absent from your callset may be absent because the reference
already carries it. Checking `ref_aa` is how you catch that before it becomes a conclusion.

Two defaults in this package follow from the same fact: `load_genotypes()` records which allele
its dosages count instead of letting REF stand in for a baseline, and `run_ihs()` / `plot_ehh()`
use `polarized = FALSE`, treating the two states as simply the two states, because ancestral
versus derived cannot be read off REF and ALT.

Reading a genome over the network costs a few seconds and it is all held in memory, so do it
once and pass the result around — `fasta =` also takes a named vector of sequences, which is
what `read_gff_cds()`'s own reader returns.

Sequence can also come from the GFF itself, when it ends with a `##FASTA` section —
`read_gff_cds()` keeps it and nothing extra needs passing:

```{r embedded-fasta}
gff <- tempfile(fileext = ".gff")
writeLines(c("##gff-version 3",
             "demo\t.\tCDS\t11\t25\t.\t+\t0\tID=c1;Parent=T.1;gene_id=T",
             "##FASTA", ">demo", "CCCCCCCCCCATGAAATTTGGGTAAC"), gff)

snp_aa_positions(data.frame(chr = "demo", pos = c(11, 14, 17, 20, 23)), read_gff_cds(gff),
                 keep = "hits", one_based_snps = TRUE)[
  , c("pos", "aa_position", "ref_codon", "ref_aa")]
```

Otherwise pass `fasta = ` a path or URL to a genome — gzipped is fine, and it reads straight
from the web like `read_gff_cds()` does. Without sequence from either source the two columns are
simply absent: a reference base cannot be inferred from an annotation. Some GFFs do carry a
translated protein in a feature attribute, but too few agree on how for it to be worth reading.

## Either attribute convention

PlasmoDB writes `Parent=PF3D7_0709000.1` alongside a `gene_id`; Ensembl writes
`Parent=transcript:PF3D7_0709000.1` and no gene id at all. `read_gff_cds()` normalises both,
deriving the gene id from the transcript when it has to. Rewriting the bundled records the
way Ensembl writes them lands on the same coordinates:

```{r conventions}
rows <- grep("^#", readLines(system.file("extdata", "pf3d7_drug_gene_cds.gff",
                                         package = "plasgenomicsutilsR")),
             invert = TRUE, value = TRUE)
ensembl_style <- sub("ID=[^;]*;Parent=([^;]*);gene_id=[^;]*.*", "Parent=transcript:\\1", rows)

sub(".*\t", "", rows[1])            # the attribute column as PlasmoDB writes it
sub(".*\t", "", ensembl_style[1])   # and as Ensembl writes it
```

```{r conventions-agree}
ens <- tempfile(fileext = ".gff")
writeLines(ensembl_style, ens)

codon76 <- function(x) aa_intervals(data.frame(transcript_id = "PF3D7_0709000.1",
                                               aa_position = 76), x, genes = NULL,
                                    one_based_output = TRUE)
both <- rbind(codon76(cds), codon76(read_gff_cds(ens)))
both$read_as <- c("PlasmoDB", "Ensembl")
both[, c("read_as", "chrom", "start", "end", "gene_id")]
```

## Watch the position base

Everything in the package is 0-based, genotype matrices included: `load_genotypes()` shifts
SNPRelate's 1-based `POS` on the way in, so a scan built from those genotypes lines up
exactly with the IBD tables and with these codon intervals. Nothing extra to do.

The argument below exists for tables that carry raw VCF positions — something you read with
`bcftools query`, a marker list typed out of a paper, or a genotype matrix from a version
before that shift. Getting it wrong is a silent one-base slide:

```{r one-based}
one_snp <- data.frame(snp_id = "Pf3D7_07_v3:403624")

# 0-based, the package convention: the second base of pfcrt codon 76
snp_aa_positions(one_snp, cds)[, c("snp_id", "aa_position", "codon_base")]

# the same number read as a VCF POS: the first base instead
snp_aa_positions(one_snp, cds, one_based_snps = TRUE)[, c("snp_id", "aa_position",
                                                          "codon_base")]
```

`annotate_snps()` takes the same argument, for the same reason. A `PopStructure` reports
which convention its own positions follow with `$positions()`.

## Putting it to work

Because `aa_intervals()` returns an ordinary interval table, it goes straight into the plots
and the interval tools. A few things this makes easy:

```{r workflow, eval = FALSE}
codons <- aa_intervals(data.frame(transcript_id = "pfdhps",
                                  aa_position   = c(437, 540, 581)), cds)

# 1. which genotyped SNPs actually are those codons?
snps <- data.frame(snp_id = colnames(ps$genotype("full")))
annotate_snps(snps, codons, keep = "hits")

# 2. mark them on a haplotype plot
plot_region_haplotypes(ps, "pfdhps", split = "region", pad = 20000, mark_snps = codons)

# 3. label a zoomed panel by residue instead of by gene
plot_ibd_locus(ibd, "pfdhps", pad = 20000, genes_for_track = codons,
               gene_label_angle = 90)

# 4. and the other direction: every coding SNP in a gene, with its residue
snp_aa_positions(snps, cds, keep = "hits") |>
  subset(gene_id == "PF3D7_0810800")
```

On a real callset that last step resolves ~28,000 genotyped SNPs to ~17,000 coding ones
across ~3,400 genes in about a second.
