---
title: "Amino acids and genomic coordinates"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Amino acids and genomic coordinates}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

Resistance markers are named by residue — *pfcrt* K76T, *pfdhps* A437G, *pfkelch13* C580Y —
while callsets, scans and every plot in this package work in genomic coordinates. Going
between the two is not arithmetic you can do in your head: it depends on the transcript's
exon structure, which strand it is on, and the CDS phase.

Two functions do it, in opposite directions:

- `aa_intervals()` — a residue → the genomic interval of its codon
- `snp_aa_positions()` — a SNP → the residue it sits on, or `NA` if it is not coding

Both read coding exons from a GFF via `read_gff_cds()`.

```{r setup}
library(plasgenomicsutilsR)
```

## Getting an annotation

`read_gff_cds()` takes a path or a URL, and will read a gzipped file straight from the web,
so there is no need to keep a copy. Two released *P. falciparum* 3D7 annotations, both of
which this package handles:

```{r sources, eval = FALSE}
# VEuPathDB / PlasmoDB — the source the bundled PF3D7_GENES was built from
cds <- read_gff_cds(paste0("https://plasmodb.org/common/downloads/Current_Release/",
                           "Pfalciparum3D7/gff/data/PlasmoDB-68_Pfalciparum3D7.gff"))

# Ensembl Protists, gzipped
cds <- read_gff_cds(paste0("https://ftp.ensemblgenomes.ebi.ac.uk/pub/protists/current/",
                           "gff3/plasmodium_falciparum/",
                           "Plasmodium_falciparum.GCA000002765v3.63.gff3.gz"))
```

Both are versioned in the filename, so bump the release number when it moves on. They name
their attributes differently — PlasmoDB writes `Parent=PF3D7_0709000.1` with a `gene_id`,
Ensembl writes `Parent=transcript:PF3D7_0709000.1` and no gene id at all — and
`read_gff_cds()` normalises both, deriving the gene id from the transcript when it has to.
They agree on the answers: *pfcrt* codon 76 comes out at `403,624-403,626` either way.

Those chunks are not run when this page is built, so the site never depends on the network.
Everything below uses a small annotation written out here instead, so the numbers can be
checked by eye.

```{r toy}
gff <- tempfile(fileext = ".gff")
writeLines(c(
  "##gff-version 3",
  # a plus-strand gene in two coding exons: 1001-1005, then 1101-1120
  "chr1\tdemo\tCDS\t1001\t1005\t.\t+\t0\tID=A-CDS1;Parent=A.1;gene_id=A",
  "chr1\tdemo\tCDS\t1101\t1120\t.\t+\t1\tID=A-CDS2;Parent=A.1;gene_id=A",
  # and a minus-strand gene in one exon: 2001-2030, read from 2030 down
  "chr2\tdemo\tCDS\t2001\t2030\t.\t-\t0\tID=B-CDS1;Parent=B.1;gene_id=B"), gff)

cds <- read_gff_cds(gff)
cds
```

## A residue to its codon

```{r aa-to-genome}
aa <- data.frame(transcript_id = "A.1", aa_position = 1:3)
aa_intervals(aa, cds, genes = NULL, one_based_output = TRUE)[
  , c("name", "chr", "start", "end", "codon_positions", "spans_intron")]
```

Codon 1 is `1001,1002,1003` and codon 2 is `1004,1005,1101` — the first exon runs out after
five bases, so codon 2 straddles the intron. When that happens `start` and `end` span the
intron too, which makes the interval 98 bp wide rather than 3, so `spans_intron` flags it and
`codon_positions` always lists the three bases themselves.

Coordinates come back **0-based half-open** by default, matching every other interval in the
package (`?"plasgenomicsutilsR-coordinates"`), which is what lets the result be used as an
interval table directly. `one_based_output = TRUE`, used above, gives the 1-based numbers you
would quote in text.

On the minus strand the transcript is read from the highest coordinate down:

```{r minus-strand}
aa_intervals(data.frame(transcript_id = "B.1", aa_position = 1:2), cds, genes = NULL,
             one_based_output = TRUE)[, c("name", "codon_positions", "strand")]
```

`transcript_id` also accepts a gene id, or a gene symbol when `genes` carries one — so with
the bundled annotation `"pfcrt"` works as well as `"PF3D7_0709000.1"`.

## A SNP to its residue

```{r genome-to-aa}
snps <- data.frame(pos = c(1001, 1004, 1101, 1050, 1121),
                   chr = "chr1")
snp_aa_positions(snps, cds, one_based_snps = TRUE)[
  , c("pos", "transcript_id", "aa_position", "codon_base", "coding")]
```

`aa_position` is **1-based**, counting the initiator methionine as 1 — the one deliberate
exception to this package's 0-based rule, at both ends: where you supply a residue and where one
is reported. Residue numbering is too settled to renumber (*pfcrt* K76T is codon 76 wherever it
is written down, and calling it 75 would be wrong in the only sense that matters). The genomic
`start`/`end` in the same table stay 0-based half-open; `codon_positions` is the one place a
1-based *position* is reported, since those are the numbers people quote. `codon_base` says which of the codon's
three bases the SNP is, in transcript orientation — so `1004` and `1101` are bases 1 and 3 of
the same codon 2. `1050` is intronic and `1121` past the CDS, so both are non-coding and come
back `NA`; `keep = "hits"` drops them instead.

On a minus-strand gene, `codon_base == 1` is the *highest* genomic coordinate of the three:

```{r minus-strand-snp}
snp_aa_positions(data.frame(chr = "chr2", pos = c(2030, 2029, 2028)), cds,
                 keep = "hits", one_based_snps = TRUE)[
  , c("pos", "aa_position", "codon_base", "strand")]
```

Being inside a codon is not the same as changing the residue: that needs the alleles and the
codon's other two bases, which neither function does.

## Watch the position base

Everything in the package is 0-based, genotype matrices included: `load_genotypes()` shifts
\pkg{SNPRelate}'s 1-based `POS` on the way in, so a scan built from those genotypes lines up
exactly with the IBD tables and with these codon intervals. Nothing extra to do.

The argument below exists for tables that carry raw VCF positions — something you read with
`bcftools query`, a marker list typed out of a paper, or a genotype matrix from a version
before that shift. Getting it wrong is a silent one-base slide, enough to turn a near-miss
into a hit:

```{r one-based}
one_snp <- data.frame(snp_id = "chr1:1004")

# 0-based, the package convention: 1004 is the second base of codon 2
snp_aa_positions(one_snp, cds)[, c("snp_id", "aa_position", "codon_base")]

# the same number read as a VCF POS: it is the first base instead
snp_aa_positions(one_snp, cds, one_based_snps = TRUE)[, c("snp_id", "aa_position",
                                                          "codon_base")]
```

`annotate_snps()` takes the same argument, for the same reason. A `PopStructure` reports which
convention its own positions follow with `$positions()`.

## Putting it to work

Because `aa_intervals()` returns an ordinary interval table, it goes straight into the plots
and the interval tools. A few things this makes easy:

```{r workflow, eval = FALSE}
codons <- aa_intervals(data.frame(transcript_id = c("pfdhps", "pfdhps", "pfdhps"),
                                  aa_position   = c(437, 540, 581)), cds)

# 1. which genotyped SNPs actually are those codons?
snps <- data.frame(snp_id = colnames(ps$genotype("full")))
annotate_snps(snps, codons, keep = "hits")

# 2. mark them on a haplotype plot
plot_region_haplotypes(ps, "pfdhps", split = "region", pad = 20000, mark_snps = codons)

# 3. label a zoomed panel by residue instead of by gene
plot_ibd_locus(ibd, "pfdhps", pad = 20000, genes_for_track = codons,
               gene_label_angle = 90)

# 4. and the other direction: every coding SNP in a gene, with its residue
snp_aa_positions(snps, cds, keep = "hits") |>
  subset(gene_id == "PF3D7_0810800")
```

On a real callset that last step resolves ~28,000 genotyped SNPs to ~17,000 coding ones across
~3,400 genes in about a second.
