---
title: "Coverage QC"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Coverage QC}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
```

```{r setup}
library(plasgenomicsutilsR)
```

The depth tables come from the Python package (`plasgenomicsutils coverage_depth_stats`
and `coverage_dropout_regions`, which read the BAMs); this package reads and plots them.

```{r, eval = FALSE}
cov <- read_coverage("coverage_by_sample.tsv.gz")
coverage_qc(cov, threshold = 10, min_mean = 5, min_breadth = 80)   # one row per sample
plot_coverage_summary(cov)     # mean vs breadth, failures labelled
plot_coverage_by_chrom(cov)    # sample x chromosome, relative to each sample's own mean

plot_coverage_dropout(read_coverage("coverage_windows.tsv.gz"),
                      genes = PF_EXAMPLE_DRUG_GENES)
```

Breadth matters more than mean depth: selective whole-genome amplification can give a
respectable average while leaving much of the genome at zero, and only the
`pct_ge_10x`-style column shows that. On one real cohort the sample that failed QC had a
mean of 122x — and a median of 33x with a quarter of the core genome under 10x.

Two things to keep straight when generating those tables. The depth engines do not agree
by definition: `mosdepth` counts **fragments** (an overlapping mate pair once) while
`pysam`/`samtools depth` count **reads**, which puts mosdepth a couple of percent lower
everywhere. Fragment depth is the better measure of independent evidence; either is fine
as long as a cohort uses one, and the `engine` column records which. And
`plot_coverage_dropout()` earns its place on the cross-sample question — a window empty in
*everyone* is not missing data, it silently reads as invariant.

## What the table looks like

The two floors are independent, and `fail_reason` says which one a sample missed:

```{r}
cov <- data.frame(
  sample = paste0("s", 1:4), chrom = "genome",
  mean   = c(60, 45, 30, 3),
  # s3 is deep on average but narrow -- amplification, not a shallow run
  pct_ge_10x = c(96, 91, 41, 7)
)
coverage_qc(cov)
```

`plot_coverage_summary()` draws the same two numbers against each other with both floors
marked, so the failures separate along whichever axis they failed on -- which is the
quickest way to tell a shallow run from an uneven one.

```{r, fig.height = 4}
plot_coverage_summary(cov)
```
