---
title: "Diversity, linkage and selection"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Diversity, linkage and selection}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(plasgenomicsutilsR)
```

Three questions about a cohort, each answered from the same genotype matrix and the same
metadata grouping: how diverse is each population, how far does linkage reach, and which
loci are under selection.

The example fixture is deliberately tiny (60 samples, 49 SNPs), so the numbers below are
shaped by that, not by biology. Everything scales to a real callset unchanged.

```{r}
ps <- example_pop_structure(umap = FALSE)
```

## Diversity

`pop_diversity()` reports, per group: nucleotide diversity, expected heterozygosity,
Watterson's theta, Tajima's D, and the haplotype / multilocus-genotype summaries.

```{r}
pop_diversity(ps, group = "country", accessible = PF3D7_CORE_REGIONS)
```

### pi and He are not the same number

This is the one thing worth being careful about. **`pi` is per accessible base pair** --
the sum of per-site heterozygosity divided by the number of callable sites. **`he`** is
that same per-site quantity averaged over the **SNPs**. They differ by a factor of
several thousand and only one of them is comparable between windows:

```{r}
d <- pop_diversity(ps, group = "country", accessible = PF3D7_CORE_REGIONS)
d[c("group", "n_snps", "n_sites", "he", "pi")]
```

Dividing by the SNP count instead of the base count is the usual mistake, and it makes
windows with different SNP density or missingness silently incomparable. `accessible`
sets the denominator; without it every base of the unit is assumed callable, which is
only right if your VCF really was called across the whole span.

### Per gene, or in windows

Per gene is the convention for *P. falciparum* Tajima's D -- coding SNPs, at least three
of them, no MAF filter:

```{r}
pop_diversity(ps, group = "country", by = "gene",
              genes = PF_EXAMPLE_DRUG_GENES)[1:4, c("group", "unit", "n_snps", "he")]
```

Windows give a genome-wide track. `step` slides the window rather than tiling it, which
is the usual way to scan Tajima's D — a fixed grid can split a signal across two windows
and dilute it in both:

```{r, fig.width = 7, fig.height = 4}
win <- pop_diversity(ps, group = "country", by = "window", window = 500000, step = 250000)
plot_diversity(win, metric = "tajima_d")
```

On a real callset `window = 5000, step = 2500` is a reasonable scanning resolution.
Overlapping windows are not independent, so count *peaks*, not windows —
[selection_peaks()] does that.

`tajima_p` tests each unit against the standard neutral model, and `tajima_percentile`
places its D within the group's own distribution. Prefer the percentile for shortlisting:
see `?tajima_d_pvalue` for why the parametric test is conservative here.

### Haploid genotypes, mixed infections

The parasite is haploid, so a heterozygous call is a mixed infection at that site rather
than a diploid genotype. By default those calls are dropped at that site (`het =
"missing"`) and every allele count is a count of samples; `het = "dosage"` splits the
call between the two alleles instead.

Tajima's D needs one sample count for its variance term, but sites differ in how many
calls they have. Rather than discard every sample with a gap, all sites are kept and the
mean number of calls is used as *n* -- `n_taj_snps` and `n_taj_samples` report what went
in. The haplotype-based columns (`hap_div`, `shannon_h`, `simpson_lambda`) do need
complete data, and there something has to give: either the samples with a gap or the sites.
Sites are the plentiful axis, so the sites with any missing call are dropped and every
sample stays in the comparison (`n_hap_snps` reports what survived). Dropping samples
instead would be hopeless at scale — at 0.05% missingness over 27,000 SNPs a sample
survives with probability ~1e-6.

Note that haplotype diversity is a *per-locus* statistic: genome-wide it saturates at 1,
because over thousands of SNPs every sample is its own haplotype. Use `by = "gene"` or
`by = "window"` for it to say anything.

## Linkage disequilibrium

Two views, and they live on different sides of the split. **r-squared decay** with
physical distance is the one genuinely quadratic statistic here -- every SNP against every
other SNP within `max_dist` -- so it runs in the Python package, where it takes about 7
seconds on a 249-sample, 28k-SNP callset:

```bash
plasgenomicsutils ld_decay --vcf cohort.bcf --meta meta.tsv --group-col region \
  --max-dist 50000 --bins 20 --output ld/decay.tsv.gz --half-decay-output ld/half.tsv
```

`read_ld_decay()` brings it back, with the half-decay distance and the scan settings
attached so the thinning behind a curve travels with it:

```{r, eval = FALSE}
ld <- read_ld_decay("ld/decay.tsv.gz", "ld/half.tsv")
attr(ld, "ld_half_decay")
plot_ld_decay(ld)
```

**`ld_index()`** is the multilocus view and stays here, being linear and quick: `Ia` and
its standardised form `rbarD`, both 0 under free recombination and rising as reproduction
becomes clonal. Compare `rbarD` between datasets, since `Ia` grows with the number of loci.

```{r}
ld_index(ps, group = "country")
```

Both are **upward-biased in small samples**, so read them next to `n_samples` -- a group
of ten will score above a group of a hundred drawn from the same population.

## Selection

Two complementary scans. Directional sweeps (drug resistance) show up in extended
haplotype homozygosity; long-term balancing selection (antigens) shows up in clustered
allele frequencies.

### Haplotypes first

\pkg{rehh} needs phased haplotypes with no missing calls, and a *P. falciparum* VCF has
neither. `parasite_haplotypes()` is the bridge, and it reports every sample and SNP it
removed -- how the haplotypes were made determines what the scan can honestly claim.

```{r}
hap <- parasite_haplotypes(ps, maf = 0.05)
hap
```

With per-sample Fws to hand (from `plasgenomicsutils calculate_fws`), gate to monoclonal
infections first, since a polyclonal infection is a mixture of haplotypes rather than one:

```{r, eval = FALSE}
fws <- readr::read_tsv("fws.tsv")
hap <- parasite_haplotypes(ps, fws = fws, min_fws = 0.95, maf = 0.03)
```

Both the mixed-call resolution and the imputation are random draws at the population
allele frequency. `seed` makes a run reproducible -- and a signal worth reporting should
survive repeating the whole thing under a different seed.

Which samples came through is worth checking before quoting a cohort size, and the same
set is usually wanted by the rest of the analysis -- see
`vignette("cohorts-and-subsets")`.

### iHS, Rsb and XP-EHH

```{r}
ihs <- run_ihs(hap, group = "country")
head(ihs, 3)
```

**How to read it.** At each SNP, iHS contrasts how far haplotype homozygosity extends from
one allele against the other. An allele that rose recently and fast still carries the long
stretch it arose on, because recombination has not had time to break it up.

Four things follow from how it is built:

* **It is a z-score.** Raw haplotype length depends heavily on allele frequency, so
  \pkg{rehh} bins SNPs by frequency and standardises within each bin. `ihs` says how
  unusual a SNP is *among SNPs at the same frequency* — which is why a warning about bins
  holding fewer than ten markers matters: the standardisation there has nothing to stand on.
* **`neg_log10_p` is a rank, not a test.** It is the normal tail of that z-score, assuming
  the bulk of the genome is neutral. Do not Bonferroni-correct it; use the empirical tail,
  `abs(ihs) > 4` or the top 1%.
* **The sign means nothing here.** With no outgroup the contrast is major-versus-minor, not
  ancestral-versus-derived. Read `abs(ihs)`.
* **Read runs, not points.** A real sweep is a *cluster* of elevated SNPs spanning tens of
  kb, since the whole haplotype is carried along. An isolated tall SNP beside ordinary
  neighbours is usually noise — which is what the Manhattan plot is for.

And one thing it cannot do: a sweep that went to **completion** leaves no minor allele to
contrast against, so iHS is blind to it. That is what the cross-population scans are for.

```{r, fig.width = 7, fig.height = 4}
plot_ihs(ihs, genes = PF_EXAMPLE_DRUG_GENES)
```

`zoom` crops any of these plots to one interval -- a chromosome, a `"chr:start-end"` range,
or a gene name from `genes` -- and names every gene in the window in a track underneath:

```{r, fig.width = 7, fig.height = 4}
plot_ihs(ihs, genes = PF_EXAMPLE_DRUG_GENES, zoom = "7", zoom_pad = 0)
```

### EHH around one SNP

A scan point says a locus looks selected; the EHH decay says what the haplotype is doing
there. `plot_ehh()` draws one curve per allele **at the focal SNP** -- so it is the
mutant-versus-reference comparison without needing the SNPs annotated -- and a sweep shows as
one allele holding EHH near 1 far past where the other has decayed:

```{r, fig.width = 7, fig.height = 3.5}
plot_ehh(hap, "pfcrt", genes = PF_EXAMPLE_DRUG_GENES, span = 30000)
```

Naming a gene leaves the choice of focal SNP to the function, which takes the most balanced one
and says so. `ehh_candidates()` shows the shortlist it chose from, most balanced first, so you
can see what was on offer and name a `chr:pos` instead:

```{r}
ehh_candidates(hap, "pfcrt", group = "country", genes = PF_EXAMPLE_DRUG_GENES)
```

`maf` is the metric that decides it. The per-group columns catch the other way a panel comes
back empty: the SNP chosen here has `maf` 0.5 yet is monomorphic *inside* both countries —
it separates them — so `plot_ehh(group = "country")` would have no curve to draw for either.
`n_groups_variable` finds the SNPs that do vary within a group.

A marker that only segregates in one part of the cohort is buried by pooling, so restrict the
haplotypes to where it varies. `subset_haplotypes()` matches metadata the way
`PopStructure$subset()` does, `column = values`, and takes several values:

```{r, fig.width = 7, fig.height = 3.5}
gh <- subset_haplotypes(hap, country = "Ghana")
plot_ehh(gh, "pfcrt", genes = PF_EXAMPLE_DRUG_GENES, span = 30000)
```

The SNP panel is left exactly as `parasite_haplotypes()` built it, so two subsets remain
comparable to each other and to the whole; a SNP that is monomorphic within the subset is
dropped when the curve is computed anyway. `print()` reports the restriction, since it changes
what every scan off that object means.

`focal` also takes a bare position or a gene name from `genes`; a gene holding several SNPs
resolves to the most balanced one and says which, since EHH from a near-singleton is a flat
line. `group` gives a panel per metadata group -- worth doing after the pooled view, because
EHH knows nothing about population structure and a group with few carriers gives a ragged
curve, which is why `min_haplotypes` drops the smallest ones. `span` crops the window, one
value or two (`c(left =, right =)`).

### The genotypes behind a signal

A scan says a locus looks selected; the genotypes say what the haplotypes actually are.
`plot_region_haplotypes()` draws one row per sample and one column per SNP over a window,
clusters the samples and puts the dendrogram beside them, with the gene track underneath:

```{r, fig.width = 8, fig.height = 6}
plot_region_haplotypes(ps, "pfcrt", pad = 20000, genes = PF_EXAMPLE_DRUG_GENES)
```

`split` blocks the rows by a metadata column and clusters *within* each block, so a haplotype
shared across a group reads as a solid band instead of being scattered by one global
ordering (this is what `ComplexHeatmap`'s `row_split` does). `mark_snps` puts a line on the
SNPs worth pointing at:

```{r, fig.width = 8, fig.height = 6}
plot_region_haplotypes(ps, "pfcrt", pad = 20000, split = "country",
                       genes = PF_EXAMPLE_DRUG_GENES, annotations = "country")
```

`annotations` adds coloured strips down the right, one column per metadata variable, each
with its own legend and taking its colours from the object's shared maps -- so a level keeps
the colour it has in the UMAP or the admixture bars:

```{r, fig.width = 8, fig.height = 6}
plot_region_haplotypes(ps, "pfdhps", pad = 20000, split = "country",
                       annotations = "country", genes = PF_EXAMPLE_DRUG_GENES)
```

`spacing` decides what the horizontal axis means, and the two answers show different things.
`"even"` (the default) gives every SNP the same width, which is how the haplotype structure
is easiest to read but says nothing about distance. `"genomic"` keeps every mark the same
width and moves it to its real coordinate, so the gaps between SNPs are what you see:

Either way a SNP is only ever drawn over the genes it really falls in. Under `"even"` the axis
counts SNPs, so a gene's box is exactly the columns it holds — its width is how many SNPs are in
it, *not* how long the gene is — and a gene with no genotyped SNP in the window has no width at
all and is left off the track with a message. Under `"genomic"` the boxes are true extents and
the marks are clipped at gene edges instead.

```{r, fig.width = 8, fig.height = 6}
plot_region_haplotypes(ps, "pfcrt", pad = 20000, split = "country",
                       spacing = "genomic", genes = PF_EXAMPLE_DRUG_GENES)
```

One thing to be sure of before reading the colours: `allele` says whether a dosage of 2 means
two alternate alleles or two reference ones. The plot asks the object, which knows when it
came from [load_genotypes()]; an object built from a bare matrix cannot say, and the two codings
are indistinguishable afterwards, so it assumes alt dosage and tells you it did.

Rsb and XP-EHH compare two populations rather than two alleles, so they see exactly those
completed sweeps:

```{r}
head(run_rsb(hap, group = "country"), 3)
```

`pairs = list(c("a", "b"))` restricts the comparison; the default is every pair.

`ihs_genes()` reduces any of these to one row per gene -- the peak SNP inside it:

```{r}
ihs_genes(ihs, genes = PF_EXAMPLE_DRUG_GENES, within = 10000, min_snps = 1)
```

### Beta

`beta_score()` looks for neighbourhoods where allele frequencies pile up around an
intermediate-frequency core -- the footprint of long-term balancing selection, and in
this parasite usually an antigen rather than a drug target.

```{r, fig.width = 7, fig.height = 3.5}
b <- beta_score(ps, group = "country", window = 300000, min_window_snps = 1)
plot_beta(b)
```

Use `window = 1000` (the default) on a real, dense callset; the fixture needs a far wider
window just to find any neighbours. `beta_genes()` summarises per gene, giving the
per-gene table these scans are usually reported as -- mean beta beside the iHS peak and
Tajima's D from `pop_diversity(by = "gene")`.
