---
title: "Filling in missed calls"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Filling in missed calls}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(plasgenomicsutilsR)
cds <- read_gff_cds(system.file("extdata", "pf3d7_drug_gene_cds.gff",
                                package = "plasgenomicsutilsR"))
ps <- example_pop_structure(umap = FALSE)
```

Filtering removes SNPs, and some of what it removes is a residue you meant to keep. Going
the other way finds them: take the codons you care about, subtract what the panel already
holds, and what is left is what was never called.

`bed_subtract()` cuts at base resolution, which is the part that matters. A codon with one
of its three bases in the panel comes back as the other two -- a residue cannot be read from
one base, so the partly covered codon is exactly the one worth re-calling.

```{r}
want <- aa_intervals(data.frame(transcript_id = c("pfcrt", "pfcrt", "pfkelch13"),
                                aa_position = c(72, 76, 580)), cds)

# a panel that happens to hold only the middle base of the K76T codon
have <- paste0("Pf3D7_07_v3:", want$start[want$aa_position == 76] + 1)

gaps <- bed_subtract(want, have)
gaps[, c("name", "aa_position", "start", "end", "piece", "width")]
```

`locs2` can be named by whatever holds the panel -- `chr:pos` ids as above, a genotype
matrix, a `load_genotypes()` list, or a `PopStructure`:

```{r}
sum(bed_subtract(want, ps)$width)   # bases of those three codons the panel lacks
```

`write_bed()` writes the result for `bcftools mpileup -R`:

```{r, eval = FALSE}
write_bed(gaps, "fill_in_snps.bed")
```

```bash
bcftools mpileup -f Pf3D7.fasta -R fill_in_snps.bed --bam-list bams.txt   | bcftools call --ploidy 2 -m -Ob -o fill_in_calls.bcf
```

It writes `chrom` in preference to `chr`. Tables here carry both -- `chr` normalised for
matching (`"7"`) and `chrom` as the source spells it (`"Pf3D7_07_v3"`) -- and a BED is read
by other tools against a real reference, so the normalised name produces a file that
matches nothing, silently.

One caveat on intron-spanning codons: `aa_intervals()` reports `[start, end)`, which for a
codon split by an intron reaches across the intron. The exact bases are in
`codon_positions`, and `spans_intron` flags the rows where the two differ.
