---
title: "IBD analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{IBD analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
have_ggplot <- requireNamespace("ggplot2", quietly = TRUE) &&
  requireNamespace("scales", quietly = TRUE)
knitr::opts_chunk$set(collapse = TRUE, comment = "#>",
                      fig.width = 8, fig.height = 5, dpi = 96, eval = have_ggplot)
library(plasgenomicsutilsR)
```

`plasgenomicsutilsR` plots identity-by-descent (IBD) results produced by the companion
Python package. You hand it the per-SNP, pairwise-group, and selection-statistic tables;
it draws genome-wide tracks, group heatmaps, tug-of-war mirrors, and drug-gene triangles.

## Generating the input tables (`plasgenomicsutils`)

The tables come from the Python IBD tools, run on your `hmmibd-rs` output, a SNP panel, a
clean BCF, and a sample → group map. The pipeline is:

```bash
# 1. Binary (pairs x SNPs) IBD matrix from hmmibd-rs blocks + a SNP panel (writes ibd_matrix.npz)
plasgenomicsutils build_ibd_matrix \
  --blocks hmmibd_rs.hmm.tsv.gz --snps panel.vcf.gz --snp-format vcf \
  --output ibd_matrix

# 2. Per-pair / per-SNP / per-group summaries (--pairwise-group-snp adds the group-pair table)
plasgenomicsutils analyze_ibd_matrix \
  --matrix ibd_matrix --meta sample_groups.tsv --group-col group \
  --pairwise-group-snp --output ibd_analysis

# 3. Global + per-group allele frequencies, single pass over the BCF
#    (writes allele_freqs.tsv.gz and group_allele_freqs.tsv.gz into the output dir)
plasgenomicsutils compute_allele_freqs \
  --bcf clean.snps.bcf --meta sample_groups.tsv --group-col group \
  --output freqs/

# 4. IBD selection statistic (XiR,s), genome-wide and per group.
#    --permute 200 adds a threshold and per-SNP p-values drawn from the data instead of
#    from a chi-square(1); see "Choosing a significance criterion" below.
plasgenomicsutils ibd_selection_statistic \
  --matrix ibd_matrix \
  --af freqs/allele_freqs.tsv.gz --af-group freqs/group_allele_freqs.tsv.gz \
  --meta sample_groups.tsv --group-col group \
  --permute 200 \
  --output ibd_selection
```

Each command's options are listed by `plasgenomicsutils <command> --help`.

## The IbdResults container

`ibd_results()` ingests those tables (paths or data frames) and precomputes the
cumulative-genome coordinate the plots share. A small **public** example dataset (five
African countries) ships with the package:

```{r ingest}
ibd <- example_ibd_results()
ibd
```

On your own data, point it at the tool outputs:

```r
ibd <- ibd_results(
  per_snp_group  = "ibd_analysis.per_snp_per_group.tsv.gz",
  pairwise_group = "ibd_analysis.per_snp_pairwise_group.tsv.gz",
  selection       = "ibd_selection.per_group.selection_stats.tsv.gz",
  threshold       = "ibd_selection.per_group.threshold.txt",
  genes           = PF_EXAMPLE_DRUG_GENES,
  blocks          = "10_maf_filter_hmm.hmm.txt",     # enables the block-based tools
  meta            = meta,                            # sample -> group
  group_col_in_meta = "region",                      # names the grouping, and orders it
  reference       = "pf3d7"
)
```

### Short IBD segments are dropped

Small IBD blocks are commonly spurious, so segments with fewer than **15 SNPs** or shorter
than **15 kb** are discarded by default when `blocks` are read — the filter is built in
rather than something you must apply beforehand. It is applied to the IBD **evidence**
only: the analyzed-sample set behind every denominator still comes from every row of the
blocks file, so a pair whose only segment was short still counts as compared, it simply
contributes no sharing.

```r
ibd_results(..., blocks = "hmm.txt")                                # >= 15 SNPs, >= 15 kb
ibd_results(..., blocks = "hmm.txt", min_block_snp = 20, min_block_kb = 25)
ibd_results(..., blocks = "hmm.txt", min_block_snp = 0, min_block_kb = 0)   # keep everything
```

Printing the object reports how many segments went, since the filter moves every downstream
fraction. The Python tools take the same `--min-block-snp` / `--min-block-kb` flags with the
same defaults, so the two sides agree.

### Ordering the groups

Groups read in a sensible order by default (a natural sort, so `site2` precedes `site10`).
To impose your own — geographic rather than alphabetical, say — make the metadata column a
factor and name it with `group_col_in_meta`; its levels then order **every** loaded table,
so facets, legends and triangle axes all agree:

```r
meta$region <- factor(meta$region,
                      levels = c("Northwest", "North", "Northeast",
                                 "East", "West", "Southwest"))
ibd <- ibd_results(..., meta = meta, group_col_in_meta = "region")
ibd$get_group_order()
```

`ibd$set_group_order(...)` changes it afterwards. Leaving out a group that the results
actually contain is an error rather than a silent `NA`; naming a group nothing uses is only
a warning.

### Dropping groups

Everything an `IbdResults` holds is summarised either per group or per group pair, so groups
are the unit to cut on. `$subset_groups()` returns a copy; `$restrict_groups()` does it in
place. `drop` is the usual form — "everything except these" — and `keep` names the survivors
instead:

```{r subset-groups}
groups <- levels(factor(ibd$get_selection()$group))
ibd$subset_groups(drop = groups[1])
```

The per-SNP, group-pair, selection and threshold tables are all filtered, and a group-pair
row survives only when **both** of its groups do — a cell against a dropped group means
nothing. With `meta` loaded, `blocks` and the analysed-sample set narrow to the surviving
groups' samples too, so `gene_ibd_overlap()`, `gene_ibd_pairs()` and `plot_ibd_network()`
follow. The group order is trimmed to whatever is left.

Narrowing groups cannot recompute a summary, so every number kept is still the one computed
over that group's full sample set — groups are dropped, never re-derived.

## Highlighting genes

The `genes` track supplies gene positions **and** display names. `PF_EXAMPLE_DRUG_GENES` is a
small bundled track of drug-resistance loci:

```{r genes}
PF_EXAMPLE_DRUG_GENES
```

Pass `highlight_genes` to pick which to mark and `label_genes = TRUE` to name them (top
panel only, at each gene's true position — labels are never nudged off-position).

## Genome-wide tracks

Per-SNP IBD along the genome, faceted by group, with two drug genes labelled:

```{r manhattan}
plot_ibd_sharing_manhattan(ibd, highlight_genes = c("pfcrt", "pfkelch13"), label_genes = TRUE)
```

The IBD selection statistic with the per-group Bonferroni threshold:

```{r selection}
plot_selection_manhattan(ibd, metric = "neg_log10_p",
                         highlight_genes = c("pfcrt", "pfkelch13"), label_genes = TRUE)
```

Every genome-wide plot can drop chromosomes (`chroms` / `skip_chr`, the rest re-laid out
contiguously):

```{r chroms}
plot_selection_manhattan(ibd, chroms = c("7", "13"),
                         highlight_genes = c("pfcrt", "pfkelch13"), label_genes = TRUE)
```

## Zooming in

`zoom` crops the same plot to one interval without changing the data behind it, so a locus
sits at the same coordinate in the zoomed and the genome-wide figure. It takes a
chromosome, a range, a gene name from the object's track, or a data frame with
chr/start/end, and every gene in the window is drawn and named in a track underneath:

```{r zoom, fig.height = 7}
plot_selection_manhattan(ibd, zoom = "7:340,000-470,000")
```

`zoom_pad` adds context around the interval -- a fraction of its span below 1, base pairs at
or above it -- which is what makes a single gene a usable window:

```{r zoom-gene, fig.height = 7}
plot_ibd_sharing_manhattan(ibd, zoom = "pfcrt", zoom_pad = 100000)
```

Two values pad the left and the right separately, in that order or named, so a window can
follow a signal that runs off one side of a gene:

```{r zoom-asym, fig.height = 7}
plot_ibd_sharing_manhattan(ibd, zoom = "pfcrt",
                           zoom_pad = c(left = 20000, right = 120000))
```

`zoom` works the same way on `plot_ibd_tugofwar()` and on the scan plots
(`plot_ihs()`, `plot_beta()`, `plot_diversity()`).

## Which genes are under selection?

`pos_selection_genes()` intersects the significant SNPs (at or above each group's
threshold) with the object's `genes` track. Because filtering can leave the peak SNP just
outside a gene, a SNP counts when it falls within `within` bp of the gene (default 2 kb):

```{r sel-genes}
pos_selection_genes(ibd)                       # object's drug-gene track, 2 kb window
```

Every metric in the selection table is reported at each gene's peak SNP (`peak_*`).
Pass `genes = PF3D7_GENES` to scan **every** gene without attaching the 5,000-gene track
to the object (which would make the Manhattan plots draw a line per gene):

```{r sel-genes-all}
pos_selection_genes(ibd, genes = PF3D7_GENES, within = 2000)
```

## Choosing a significance criterion

`ibd_selection_statistic` reports up to four cutoffs, and `get_thresholds()` shows them
beside the diagnostic that says which to use:

```{r sel-thresholds}
ibd$get_thresholds()
```

Two read their p-values off a chi-square(1): **Bonferroni** (family-wise) and
**Benjamini-Hochberg** (a stated share of the calls may be false, so it calls more). Two
come from `--permute`, which rebuilds the null by sliding each pair's IBD segments to a
random circular offset — preserving that pair's total sharing, segment count, segment
lengths and along-genome autocorrelation, and destroying only the alignment of segments
*between* pairs: **`permutation`** (family-wise, from the per-replicate genome-wide maxima)
and **`empirical`** (Benjamini-Hochberg over the permutation's own per-SNP p-values).

`lambda_gc` decides between them. It is 1 when the chi-square(1) reference fits, and on real
*P. falciparum* data it comes out near 0.1: the within-MAF-bin standardisation fixes the
z-scores' mean and variance and nothing about their shape. Where it is far from 1, prefer the
permutation lines — `pval` and `q_value` are still fine for *ranking* SNPs, since every step
from `z_score` to `neg_log10_p` is monotone, but they are not probabilities.

Draw any one of them, or `"all"` for every kind the run wrote (chi-square lines warm,
permutation lines cool). The shipped example comes from a run without `--permute`, so only
the Bonferroni line is available here:

```{r sel-thresholds-plot, fig.height = 7}
plot_selection_manhattan(ibd, draw_threshold = "all", label_genes = TRUE)
```

Then merge whichever criterion you settle on into loci, rather than reporting SNP counts —
a sweep spans many SNPs, so a SNP count largely measures panel density:

```{r sel-peaks-criteria}
selection_peaks(ibd, criterion = "bonferroni", genes = PF3D7_GENES, min_snps = 2)
```

With a `--permute` run the same call takes `criterion = "permutation"` or `"empirical"`,
and `draw_threshold` the matching names:

```{r sel-peaks-perm, eval = FALSE}
selection_peaks(ibd, criterion = "permutation", genes = PF3D7_GENES, min_snps = 2)
selection_peaks(ibd, criterion = "empirical", genes = PF3D7_GENES, min_snps = 2)
```

## Tug-of-war

Selection hangs from the top, IBD rises from the bottom, sharing one colour-coded axis.
`scale = "common"` keeps panels comparable; `scale = "free"` lets each group use its own
maximum:

```{r tugofwar, fig.height = 7}
plot_ibd_tugofwar(ibd, highlight_genes = c("pfcrt", "pfdhps"), label_genes = TRUE)
```

`top` replaces the upper track with any per-SNP scan, which is how you ask whether the IBD
signal coincides with a haplotype signal rather than with the statistic derived from the
same segments. The scan needs a `group` column matching the IBD groups; two-population
scans (`run_rsb()`, `run_xpehh()`) are indexed by population pair and cannot be mirrored
against per-group IBD.

```{r tugofwar-ihs, eval = FALSE}
# hap <- parasite_haplotypes(ps, maf = 0.05)
plot_ibd_tugofwar(ibd, top = run_ihs(hap, group = "region"), top_label = "iHS")
```

## One locus in detail

`plot_ibd_locus()` puts the two signals on two axes over a single window: IBD sharing as a
step curve on the left, the scan as points on the right, genes underneath. The axes carry
different units, so read shapes and positions rather than comparing heights.

```{r locus, fig.height = 7}
plot_ibd_locus(ibd, "pfcrt", pad = 50000)
```

`size_by` and `shape_by` map extra scan columns onto the points when the scan has them --
`shape_by` picks up a mutation-class column (`mutation_type`, `mutation_class`, `effect`,
`consequence`) on its own, and `NA` turns either off.

### Showing every gene in the window

The gene track and the marks inside the panel can come from different tables.
`genes_for_track` fills the track from a full annotation so the whole neighbourhood is
visible, while the object's own short track still decides what gets marked -- without it the
only way to see the neighbours is to attach the whole annotation, which draws a reference
line per gene. Systematic ids are much wider than the genes they sit under, so
`gene_label_angle` turns them out of each other's way:

```{r locus-all-genes, fig.height = 7}
plot_ibd_locus(ibd, "pfcrt", pad = 50000,
               genes_for_track = PF3D7_GENES, gene_label_angle = 45)
```

Both arguments work on any zoomed plot, including the scan plots.

## Region-by-group sharing

IBD between group pairs along the genome; `trans = "log2"` and a single-hue ramp keep it
readable when most values are near zero:

```{r heatmap, fig.height = 6}
plot_ibd_pairwise_group_heatmap(ibd, trans = "log2")
```

Triangles ask whether a gene (or a specific locus) is itself shared:

```r
plot_pairwise_ibd_for_genes(ibd)                              # one facet per gene
plot_pairwise_ibd_for_genes(ibd, snps = "Pf3D7_07_v3:403222") # a single locus
plot_pairwise_ibd_for_genes(ibd, individual = TRUE)           # a list, one plot per feature

# pages from `individual = TRUE` are colour-comparable only on a shared scale
save_plot("triangles.pdf",
          plot_pairwise_ibd_for_genes(ibd, individual = TRUE, limits = "shared"))
```

With IBD `blocks` loaded, a gene's cell is the fraction of pairs whose IBD **segment
overlaps the gene** — so a segment spanning the gene counts even where no SNP was
genotyped inside it. Without blocks it falls back to aggregating the SNPs inside the gene
(`agg = "mean"`/`"median"`/`"max"`). Gene spans are CDS, which can be short relative to a
sparse panel, so `within` widens the window on either path:

```r
plot_pairwise_ibd_for_genes(ibd, within = 10000)   # count SNPs / blocks within 10 kb
```

## Which pairs share a gene

`gene_ibd_overlap()` gives the fraction per group pair; `gene_ibd_pairs()` gives the
adjacency list underneath it — one row per sample pair × IBD block × gene, with how much of
the gene that block covers. Pairs with no IBD over a gene are simply absent.

```r
ov <- gene_ibd_overlap(ibd, genes = c("pfcrt", "pfdhps"))   # fractions per group pair

pairs <- gene_ibd_pairs(ibd, genes = c("pfcrt", "pfdhps"))
subset(pairs, coverage == "complete")     # pairs sharing the whole gene
table(pairs$gene, pairs$coverage)         # complete vs partial, per gene
```

Columns: `sample1`/`sample2` (order-normalised), `chr`, `block_start`/`block_end`, `gene`,
`gene_id`, `gene_start`/`gene_end`, `coverage` (`complete`/`partial`),
`covered_start`/`covered_end` (the gene's own bounds when complete), `covered_bp` and
`percent_covered`. The Python `plasgenomicsutils ibd_gene_pairs` writes the same table as
a TSV.

## Sample-level networks

Each node is a sample, each edge a pair sharing an IBD block over the gene or locus. Colour
and shape are independent metadata columns, so two variables can be read off one plot:

```r
plot_ibd_network(ibd, gene = "pfcrt", color_group = "region")
plot_ibd_network(ibd, gene = "pfcrt", color_group = "region", shape_group = "collection_year")

# `colors` / `shapes` take a NAMED vector (mapped by name, may be partial) or an unnamed
# one (positional, in the column's level order)
plot_ibd_network(ibd, gene = "pfcrt", color_group = "region",
                 colors = c(North = "#1f78b4", East = "#33a02c"))
```

An edge means "this pair is IBD here", and `sharing` sets how strictly: `"overlap"`
(default) accepts a segment touching the interval anywhere, `"complete"` demands one
spanning the whole gene/locus. Which you want depends on the question — *any* shared
ancestry at the locus, or the whole locus inherited together:

```r
plot_ibd_network(ibd, gene = "pfdhps", color_group = "region")                       # overlap
plot_ibd_network(ibd, gene = "pfdhps", color_group = "region", sharing = "complete")
```

The two can differ a lot: on one real cohort 90% of the pairs sharing `pfcrt` share all of
it, but only 45% for `pfdhps`, whose IBD segments are more fragmented. The subtitle records
which criterion drew the graph, and `gene_ibd_pairs()` shows per pair which segments are
`"complete"` versus `"partial"`.

Densely inter-connected groups collapse into an unreadable blob under a plain force
layout, so edge attraction is weighted by `(1 - J)^spread`, `J` being the Jaccard overlap
of two samples' IBD neighbourhoods: redundant edges inside a near-clique pull only weakly
and the group opens into a disc, while edges bridging distinct groups keep full pull.
`spread = 1.5` by default; raise it to loosen the densest groups further, or use
`spread = 0` for plain `fr`.

## Saving

`save_plot()` sizes the canvas to the drawing. Plots with a locked panel shape — the
networks and the gene triangles both use `coord_fixed()` — only fill a canvas of one
particular aspect, and on any other shape the remainder shows up as blank margin. Give one
dimension and the other is computed from the built plot's panel ratio and the space its
titles, legends and margins actually need; give neither and both are worked out; give both
and they are used as-is. `fit = FALSE` opts out.

```r
save_plot("network.pdf", p)                 # both worked out, no leftover margin
save_plot("network.pdf", p, width = 12)     # height follows from the contents
save_plot("network.pdf", p, height = 6)     # width follows instead
save_plot("network.pdf", p, width = 9, height = 9)   # exactly as asked
```

```r
save_plot("ibd_tugofwar.pdf", plot_ibd_tugofwar(ibd))   # auto-sized; cairo PDF by default
save_plot("triangles.pdf", plot_pairwise_ibd_for_genes(ibd, individual = TRUE))  # multi-page
```
