---
title: "IBD analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{IBD analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
have_ggplot <- requireNamespace("ggplot2", quietly = TRUE) &&
  requireNamespace("scales", quietly = TRUE)
knitr::opts_chunk$set(collapse = TRUE, comment = "#>",
                      fig.width = 8, fig.height = 5, dpi = 96, eval = have_ggplot)
library(plasgenomicsutilsR)
```

`plasgenomicsutilsR` plots identity-by-descent (IBD) results produced by the companion
Python package. You hand it the per-SNP, pairwise-group, and selection-statistic tables;
it draws genome-wide tracks, group heatmaps, tug-of-war mirrors, and drug-gene triangles.

## Generating the input tables (`plasgenomicsutils`)

The tables come from the Python IBD tools, run on your `hmmibd-rs` output, a SNP panel, a
clean BCF, and a sample → group map. The pipeline is:

```bash
# 1. Binary (pairs x SNPs) IBD matrix from hmmibd-rs blocks + a SNP panel (writes ibd_matrix.npz)
plasgenomicsutils build_ibd_matrix \
  --blocks hmmibd_rs.hmm.tsv.gz --snps panel.vcf.gz --snp-format vcf \
  --output ibd_matrix

# 2. Per-pair / per-SNP / per-group summaries (--pairwise-group-snp adds the group-pair table)
plasgenomicsutils analyze_ibd_matrix \
  --matrix ibd_matrix --meta sample_groups.tsv --group-col group \
  --pairwise-group-snp --output ibd_analysis

# 3. Global + per-group allele frequencies, single pass over the BCF
#    (writes allele_freqs.tsv.gz and group_allele_freqs.tsv.gz into the output dir)
plasgenomicsutils compute_allele_freqs \
  --bcf clean.snps.bcf --meta sample_groups.tsv --group-col group \
  --output freqs/

# 4. IBD selection statistic (XiR,s), genome-wide and per group
plasgenomicsutils ibd_selection_statistic \
  --matrix ibd_matrix \
  --af freqs/allele_freqs.tsv.gz --af-group freqs/group_allele_freqs.tsv.gz \
  --meta sample_groups.tsv --group-col group \
  --output ibd_selection
```

Each command's options are listed by `plasgenomicsutils <command> --help`.

## The IbdResults container

`ibd_results()` ingests those tables (paths or data frames) and precomputes the
cumulative-genome coordinate the plots share. A small **public** example dataset (five
African countries) ships with the package:

```{r ingest}
ibd <- example_ibd_results()
ibd
```

On your own data, point it at the tool outputs:

```r
ibd <- ibd_results(
  per_snp_group  = "ibd_analysis.per_snp_per_group.tsv.gz",
  pairwise_group = "ibd_analysis.per_snp_pairwise_group.tsv.gz",
  selection       = "ibd_selection.per_group.selection_stats.tsv.gz",
  threshold       = "ibd_selection.per_group.threshold.txt",
  genes           = PF_EXAMPLE_DRUG_GENES,
  reference       = "pf3d7"
)
```

## Highlighting genes

The `genes` track supplies gene positions **and** display names. `PF_EXAMPLE_DRUG_GENES` is a
small bundled track of drug-resistance loci:

```{r genes}
PF_EXAMPLE_DRUG_GENES
```

Pass `highlight_genes` to pick which to mark and `label_genes = TRUE` to name them (top
panel only, at each gene's true position — labels are never nudged off-position).

## Genome-wide tracks

Per-SNP IBD along the genome, faceted by group, with two drug genes labelled:

```{r manhattan}
plot_ibd_sharing_manhattan(ibd, highlight_genes = c("pfcrt", "pfkelch13"), label_genes = TRUE)
```

The IBD selection statistic with the per-group Bonferroni threshold:

```{r selection}
plot_selection_manhattan(ibd, metric = "neg_log10_p",
                         highlight_genes = c("pfcrt", "pfkelch13"), label_genes = TRUE)
```

Every genome-wide plot can drop chromosomes (`chroms` / `skip_chr`, the rest re-laid out
contiguously):

```{r zoom}
plot_selection_manhattan(ibd, chroms = c("7", "13"),
                         highlight_genes = c("pfcrt", "pfkelch13"), label_genes = TRUE)
```

## Which genes are under selection?

`pos_selection_genes()` intersects the significant SNPs (at or above each group's
threshold) with the object's `genes` track. Because filtering can leave the peak SNP just
outside a gene, a SNP counts when it falls within `within` bp of the gene (default 2 kb):

```{r sel-genes}
pos_selection_genes(ibd)                       # object's drug-gene track, 2 kb window
```

Every metric in the selection table is reported at each gene's peak SNP (`peak_*`).
Pass `genes = PF3D7_GENES` to scan **every** gene without attaching the 5,000-gene track
to the object (which would make the Manhattan plots draw a line per gene):

```{r sel-genes-all}
pos_selection_genes(ibd, genes = PF3D7_GENES, within = 2000)
```

## Tug-of-war

Selection hangs from the top, IBD rises from the bottom, sharing one colour-coded axis.
`scale = "common"` keeps panels comparable; `scale = "free"` lets each group use its own
maximum:

```{r tugofwar, fig.height = 7}
plot_ibd_tugofwar(ibd, highlight_genes = c("pfcrt", "pfdhps"), label_genes = TRUE)
```

## Region-by-group sharing

IBD between group pairs along the genome; `trans = "log2"` and a single-hue ramp keep it
readable when most values are near zero:

```{r heatmap, fig.height = 6}
plot_ibd_pairwise_group_heatmap(ibd, trans = "log2")
```

Triangles ask whether a gene (or a specific locus) is itself shared. A gene's cell
aggregates **all** SNPs strictly inside it (`agg = "mean"`/`"median"`/`"max"`):

```r
plot_pairwise_ibd_for_genes(ibd)                              # one facet per gene
plot_pairwise_ibd_for_genes(ibd, snps = "Pf3D7_07_v3:403222") # a single locus
plot_pairwise_ibd_for_genes(ibd, individual = TRUE)           # a list, one plot per feature
```

## Saving

```r
save_plot("ibd_tugofwar.pdf", plot_ibd_tugofwar(ibd))   # auto-sized; cairo PDF by default
```
