---
title: "IBD analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{IBD analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
have_ggplot <- requireNamespace("ggplot2", quietly = TRUE) &&
  requireNamespace("scales", quietly = TRUE)
knitr::opts_chunk$set(collapse = TRUE, comment = "#>",
                      fig.width = 8, fig.height = 5, dpi = 96, eval = have_ggplot)
library(plasgenomicsutilsR)
```

`plasgenomicsutilsR` plots identity-by-descent (IBD) results produced by the companion
Python package. You hand it the per-SNP, pairwise-region, and selection-statistic tables;
it draws genome-wide tracks, region heatmaps, tug-of-war mirrors, and drug-gene triangles.

## Generating the input tables (`plasgenomicsutils`)

The tables come from the Python IBD tools, run on your `hmmibd-rs` output, a SNP panel, a
clean BCF, and a sample → region map. The pipeline is:

```bash
# 1. Binary (pairs x SNPs) IBD matrix from hmmibd-rs blocks + a SNP panel
plasgenomicsutils build_ibd_matrix \
  --blocks hmmibd_rs.hmm.tsv.gz --snps panel.vcf.gz --snp-format vcf \
  --output ibd_matrix.npz

# 2. Per-SNP-per-region and pairwise-region IBD summaries (needs the region map)
plasgenomicsutils analyze_ibd_matrix \
  --matrix ibd_matrix.npz --meta sample_regions.tsv --region-col region \
  --output ibd_analysis

# 3. Global + per-region allele frequencies, single pass over the BCF
plasgenomicsutils compute_allele_freqs \
  --bcf clean.snps.bcf --meta sample_regions.tsv --region-col region \
  --output allele_freqs

# 4. IBD selection statistic (XiR,s), genome-wide and per region
plasgenomicsutils ibd_selection_statistic \
  --matrix ibd_matrix.npz \
  --af allele_freqs.global.tsv.gz --af-region allele_freqs.per_region.tsv.gz \
  --meta sample_regions.tsv --region-col region \
  --output ibd_selection
```

Each command's options are listed by `plasgenomicsutils <command> --help`.

## The IbdResults container

`ibd_results()` ingests those tables (paths or data frames) and precomputes the
cumulative-genome coordinate the plots share. A small **public** example dataset (five
African countries) ships with the package:

```{r ingest}
ibd <- example_ibd_results()
ibd
```

On your own data, point it at the tool outputs:

```r
ibd <- ibd_results(
  per_snp_region  = "ibd_analysis.per_snp_per_region.tsv.gz",
  pairwise_region = "ibd_analysis.per_snp_pairwise_region.tsv.gz",
  selection       = "ibd_selection.per_region.selection_stats.tsv.gz",
  threshold       = "ibd_selection.per_region.threshold.txt",
  genes           = EXAMPLE_DRUG_GENES,
  reference       = "pf3d7"
)
```

## Highlighting genes

The `genes` track supplies gene positions **and** display names. `EXAMPLE_DRUG_GENES` is a
small bundled track of drug-resistance loci:

```{r genes}
EXAMPLE_DRUG_GENES
```

Pass `highlight_genes` to pick which to mark and `label_genes = TRUE` to name them (top
panel only, at each gene's true position — labels are never nudged off-position).

## Genome-wide tracks

Per-SNP IBD along the genome, faceted by region, with two drug genes labelled:

```{r manhattan}
plot_ibd_manhattan(ibd, highlight_genes = c("crt", "kelch13"), label_genes = TRUE)
```

The IBD selection statistic with the per-region Bonferroni threshold:

```{r selection}
plot_selection_manhattan(ibd, metric = "neg_log10_p",
                         highlight_genes = c("crt", "kelch13"), label_genes = TRUE)
```

Every genome-wide plot can drop chromosomes (`chroms` / `skip_chr`, the rest re-laid out
contiguously):

```{r zoom}
plot_selection_manhattan(ibd, chroms = c("7", "13"),
                         highlight_genes = c("crt", "kelch13"), label_genes = TRUE)
```

## Tug-of-war

Selection hangs from the top, IBD rises from the bottom, sharing one colour-coded axis.
`scale = "common"` keeps panels comparable; `scale = "free"` lets each region use its own
maximum:

```{r tugofwar, fig.height = 7}
plot_ibd_tugofwar(ibd, highlight_genes = c("crt", "dhps"), label_genes = TRUE)
```

## Region-by-region sharing

IBD between region pairs along the genome; `trans = "log2"` and a single-hue ramp keep it
readable when most values are near zero:

```{r heatmap, fig.height = 6}
plot_ibd_region_heatmap(ibd, trans = "log2")
```

Triangles ask whether a gene (or a specific locus) is itself shared. A gene's cell
aggregates **all** SNPs strictly inside it (`agg = "mean"`/`"median"`/`"max"`):

```r
plot_drug_gene_triangles(ibd)                              # one facet per gene
plot_drug_gene_triangles(ibd, snps = "Pf3D7_07_v3:403222") # a single locus
plot_drug_gene_triangles(ibd, individual = TRUE)           # a list, one plot per feature
```

## Saving

```r
save_plot("ibd_tugofwar.pdf", plot_ibd_tugofwar(ibd))   # auto-sized; cairo PDF by default
```
