---
title: "Visualizing IBD results with plasgenomicsutilsR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Visualizing IBD results with plasgenomicsutilsR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
have_ggplot <- requireNamespace("ggplot2", quietly = TRUE) &&
  requireNamespace("scales", quietly = TRUE)
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  fig.width = 8, fig.height = 5, dpi = 96,
  eval = have_ggplot
)
library(plasgenomicsutilsR)
```

`plasgenomicsutilsR` plots identity-by-descent (IBD) results produced by the companion
Python package `plasgenomicsutils`. You hand it the per-SNP, pairwise-region, and
selection-statistic tables; it draws genome-wide Manhattan tracks, region-by-region
heatmaps, "tug-of-war" mirrors, and drug-gene triangles.

## The IbdResults container

`ibd_results()` ingests the tables (file paths or data frames) and precomputes the
cumulative-genome coordinate the plots share. A small **public** example dataset (five
African countries) ships with the package:

```{r}
ibd <- example_ibd_results()
ibd
```

On your own data, point it at the outputs of the `plasgenomicsutils ibd` tools:

```r
ibd <- ibd_results(
  per_snp_region  = "ibd_analysis.per_snp_per_region.tsv.gz",
  pairwise_region = "ibd_analysis.per_snp_pairwise_region.tsv.gz",
  selection       = "ibd_selection_analysis.per_region.selection_stats.tsv.gz",
  threshold       = "ibd_selection_analysis.per_region.threshold.txt",
  genes           = EXAMPLE_DRUG_GENES,
  reference       = "pf3d7"
)
```

## Genome-wide tracks

Per-SNP IBD along the genome, faceted by region:

```{r manhattan}
plot_ibd_manhattan(ibd)
```

The IBD selection statistic, with the per-region Bonferroni threshold:

```{r selection}
plot_selection_manhattan(ibd, metric = "neg_log10_p")
```

Every genome-wide plot can drop chromosomes (`chroms` / `skip_chr`, the rest re-laid
out contiguously) and highlight genes from the `genes` track. Gene names are labelled
only on the top panel and never moved off their true position:

```{r zoom}
plot_selection_manhattan(ibd, chroms = c("7", "8"),
                         highlight_genes = c("crt", "dhps"), label_genes = TRUE)
```

## Tug-of-war

Selection hangs from the top, IBD rises from the bottom, sharing one colour-coded
axis. `scale = "common"` keeps panels comparable; `scale = "free"` lets each region
use its own maximum:

```{r tugofwar, fig.height = 7}
plot_ibd_tugofwar(ibd, highlight_genes = "crt", label_genes = TRUE)
```

## Region-by-region sharing

The heatmap shows IBD between region pairs along the genome; `trans = "log2"` and a
single-hue ramp keep it readable when most values are near zero:

```{r heatmap, fig.height = 6}
plot_ibd_region_heatmap(ibd, trans = "log2")
```

Triangles ask whether a gene (or a specific locus) is itself shared. A gene's cell
aggregates **all** SNPs strictly inside it (`agg = "mean"`/`"median"`/`"max"`):

```r
plot_drug_gene_triangles(ibd)                      # one facet per gene
plot_drug_gene_triangles(ibd, snps = "Pf3D7_07_v3:403222")   # a single locus
plot_drug_gene_triangles(ibd, individual = TRUE)   # a list, one plot per feature
```

## Saving

`save_plot()` sizes the output from how much was drawn (regions × chromosomes) and,
for PDF, prefers the cairo device for reliable fonts (falling back where cairo is
unavailable, e.g. Windows):

```r
save_plot("ibd_tugofwar.pdf", plot_ibd_tugofwar(ibd))
```

## Population structure

`PopStructure` is an R6 workspace that bundles a genotype matrix, its PCA (the full
`prcomp`), an optional UMAP embedding, per-sample metadata, a **shared colour map**, and
an sNMF admixture fit — so the three plots colour and order consistently, and the whole
object can be `saveRDS`-ed and re-plotted without recomputing. `example_pop_structure()`
loads a small **public** genotype matrix (60 Pf7 samples, 30 Ghana + 30 Cambodia):

```{r popstructure, eval = requireNamespace("uwot", quietly = TRUE) && have_ggplot, fig.width = 6, fig.height = 4.5}
ps <- example_pop_structure()      # PCA + UMAP on real Ghana/Cambodia genotypes
ps
ps$plot_pca(colour = "country")
ps$plot_umap(colour = "country")
```

Feed the UMAP a chosen number of PCs, or a **fraction of variance** to capture — `pop_structure()`
runs the PCA first and picks that many components (`n_pcs_for_variance()`):

```r
ps$run_umap(pca_components = 0.5)   # PCs covering 50% of the variance
west <- ps$subset(country = "Ghana")  # sub-select samples (or by metadata) for output
```

On your own data, build from a matrix (or an LD-pruned VCF via `run_ld_prune()`) and
attach metadata; colours are auto-assigned per column and reused across plots:

```r
ps <- PopStructure$new(genotypes, meta = sample_meta)   # sample, region, country, ...
ps$run_umap(pca_components = 10)
```

For admixture, `run_snmf()` fits sNMF over a range of K (via LEA) — cached and quiet by
default (it is slow and noisy), `best_k()` picks K by cross-entropy, and the admixture
bars can be ordered once (`admixture_order()`) and reused across K, with a group strip
whose colours **match the UMAP**:

```r
ps$run_snmf(K = 1:8)                        # cached: re-runs are instant; no console spam
ord <- admixture_order(ps$q(ps$best_k()), meta = ps$get_meta(), group = "country")
ps$plot_admixture(K = ps$best_k(), group = "country", sample_order = ord)
```

### One combined figure

`plot_structure_figure()` draws the UMAP and the admixture as **one** figure sharing a
single theme (matching fonts), one region colour map (UMAP points match the admixture
colour strips), and collected legends. The admixture is faceted by region — a colour
strip, no text label — over one or more rows you specify, and the UMAP sits above
(`"vertical"`) or beside (`"horizontal"`) it. A richer public example (258 East-African
samples across DRC / Kenya / Tanzania / Uganda sites) shows it off:

```r
ps <- example_pop_structure("africa")   # UMAP baked in
ps$run_snmf(K = 1:9)

# fix the order used everywhere (legend, facets, strips) once
ps$set_levels("site", c("DRC", "Uganda_North", "Uganda_North-East", "Uganda_East",
                        "Kenya_East", "Kenya_West", "Tanzania_East", "Tanzania_West"))

plot_structure_figure(ps, group = "site", K = "best_k",  # or a specific K, e.g. K = 9
  orientation = "vertical",
  rows = list(c("DRC", "Uganda_North", "Uganda_North-East", "Uganda_East"),
              c("Kenya_East", "Kenya_West", "Tanzania_East", "Tanzania_West")),
  file = "figure2.pdf")                  # auto-sized; width/height also returned as attributes
```

### Saving the workspace

A fitted `PopStructure` holds the genotypes, PCA, UMAP, metadata, colours, and sNMF fit —
`ps$save("popstruct.rds")` and `load_pop_structure("popstruct.rds")` persist and restore
the whole thing (for the admixture to survive a reload, fit it with a persistent
`cache_dir`).

### Which SNPs differentiate regions? (Jost's D, Gst, G′st, Fst)

`pop_diff()` computes a per-SNP differentiation statistic for every pair of metadata
groups from the **full, unpruned** genotypes (LD-pruning throws away the differentiating
SNPs you want here): Jost's D (`statistic = "jost_d"`, and the `jost_d()` shortcut), Nei's
Gst, Hedrick's standardized G′st, and Hudson's Fst. `plot_diff_heatmap()` draws a triangle
heatmap with the legend tucked in the empty corner, an optional clustering **dendrogram**,
and a metadata **annotation** strip; `top_differentiating_snps()` returns the most
differentiating markers, walking the pairwise comparisons round-robin so each pair
contributes its top SNPs (this is exactly how the `"africa"` fixture is built).

```r
ps <- example_pop_structure("africa")
pd <- ps$pop_diff(group = "site", statistic = "jost_d")   # per-SNP D for all site pairs

# triangle heatmap: clustering dendrogram + two annotation strips (custom country colours)
ps$plot_diff_heatmap(group = "site", stat = "top_mean",
                     annotate = c("country", "region"),
                     annotate_colours = list(country = c(DRC = "#4C72B0", Kenya = "#DD8452",
                                                         Tanzania = "#55A868", Uganda = "#C44E52")))

# every statistic and summary side by side, one row per site pair
ps$pop_diff_table(group = "site")     # jost_d/gst/gst_hedrick/fst x mean/top_mean/max
markers <- top_differentiating_snps(pd, 2000)
```

**Reading near-zero measures.** Across the whole *P. falciparum* genome most SNPs barely
differ between African populations, so a genome-wide **mean** differentiation looks close
to zero even where real structure exists in a minority of loci. Options that surface it:
summarise with `stat = "top_mean"` (mean of the top few % of SNPs per pair) or `"max"`
rather than `"mean"`; prefer the standardized **G′st** or **Jost's D** over Nei's Gst
(which is the most deflated when within-group diversity is high); apply a `trans = "sqrt"`
fill; and remember the differentiation lives in the *loci* — `top_differentiating_snps()`
and a per-SNP genome scan are the right lens, not the genome-wide average.
