---
title: "Visualizing IBD results with plasgenomicsutilsR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Visualizing IBD results with plasgenomicsutilsR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
have_ggplot <- requireNamespace("ggplot2", quietly = TRUE) &&
  requireNamespace("scales", quietly = TRUE)
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  fig.width = 8, fig.height = 5, dpi = 96,
  eval = have_ggplot
)
library(plasgenomicsutilsR)
```

`plasgenomicsutilsR` plots identity-by-descent (IBD) results produced by the companion
Python package `plasgenomicsutils`. You hand it the per-SNP, pairwise-region, and
selection-statistic tables; it draws genome-wide Manhattan tracks, region-by-region
heatmaps, "tug-of-war" mirrors, and drug-gene triangles.

## The IbdResults container

`ibd_results()` ingests the tables (file paths or data frames) and precomputes the
cumulative-genome coordinate the plots share. A small **public** example dataset (five
African countries) ships with the package:

```{r}
ibd <- example_ibd_results()
ibd
```

On your own data, point it at the outputs of the `plasgenomicsutils ibd` tools:

```r
ibd <- ibd_results(
  per_snp_region  = "ibd_analysis.per_snp_per_region.tsv.gz",
  pairwise_region = "ibd_analysis.per_snp_pairwise_region.tsv.gz",
  selection       = "ibd_selection_analysis.per_region.selection_stats.tsv.gz",
  threshold       = "ibd_selection_analysis.per_region.threshold.txt",
  genes           = EXAMPLE_DRUG_GENES,
  reference       = "pf3d7"
)
```

## Genome-wide tracks

Per-SNP IBD along the genome, faceted by region:

```{r manhattan}
plot_ibd_manhattan(ibd)
```

The IBD selection statistic, with the per-region Bonferroni threshold:

```{r selection}
plot_selection_manhattan(ibd, metric = "neg_log10_p")
```

Every genome-wide plot can drop chromosomes (`chroms` / `skip_chr`, the rest re-laid
out contiguously) and highlight genes from the `genes` track. Gene names are labelled
only on the top panel and never moved off their true position:

```{r zoom}
plot_selection_manhattan(ibd, chroms = c("7", "8"),
                         highlight_genes = c("crt", "dhps"), label_genes = TRUE)
```

## Tug-of-war

Selection hangs from the top, IBD rises from the bottom, sharing one colour-coded
axis. `scale = "common"` keeps panels comparable; `scale = "free"` lets each region
use its own maximum:

```{r tugofwar, fig.height = 7}
plot_ibd_tugofwar(ibd, highlight_genes = "crt", label_genes = TRUE)
```

## Region-by-region sharing

The heatmap shows IBD between region pairs along the genome; `trans = "log2"` and a
single-hue ramp keep it readable when most values are near zero:

```{r heatmap, fig.height = 6}
plot_ibd_region_heatmap(ibd, trans = "log2")
```

Triangles ask whether a gene (or a specific locus) is itself shared. A gene's cell
aggregates **all** SNPs strictly inside it (`agg = "mean"`/`"median"`/`"max"`):

```r
plot_drug_gene_triangles(ibd)                      # one facet per gene
plot_drug_gene_triangles(ibd, snps = "Pf3D7_07_v3:403222")   # a single locus
plot_drug_gene_triangles(ibd, individual = TRUE)   # a list, one plot per feature
```

## Saving

`save_plot()` sizes the output from how much was drawn (regions × chromosomes) and,
for PDF, prefers the cairo device for reliable fonts (falling back where cairo is
unavailable, e.g. Windows):

```r
save_plot("ibd_tugofwar.pdf", plot_ibd_tugofwar(ibd))
```
