---
title: "Avoiding tandem repeats"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Avoiding tandem repeats}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(plasgenomicsutilsR)
```

A polymerase slips on a short tandem repeat, and the shorter the repeating unit the shorter
a run it takes: a dozen `A`s is enough, a dozen bases of `AT` too, while a trinucleotide has
to run to about twenty before it matters and anything at all is trouble from fifty. Designing
targets -- amplicons, probes, windows to call in -- means knowing where those runs are and
keeping a few bases clear of them. This article goes from a repeat finder's output to a BED
of targets with the repeats cut out, two ways: from whole genes, and from their exons.

## The repeats

`pf3d7_tandem_repeats()` is every short tandem repeat in the Pf3D7 reference, found once by
Tandem Repeats Finder and an exhaustive search for perfect repeats of units up to 10 bp
(`?pf3d7_tandem_repeats` has the exact commands), and bundled so it never has to be run
again. `tandem_repeats()` reads the same kind of BED for any other genome: only the first
four fields are used, and the name has to carry the unit and copy number as
`UNIT_xCOPIES`, with or without a `chrom-start-end__` in front.

```{r}
tr <- pf3d7_tandem_repeats()
nrow(tr)
tr[1:4, c("chrom", "start", "end", "width", "repeat_unit", "period", "copies")]
```

`period` is the length of the *true* repeating unit, whatever spelling the finder chose --
`AT`, `ATAT` and `ATATAT` all describe a dinucleotide run and all get `period = 2`. That is
the number the slippage rules key on:

```{r}
table(pmin(tr$period, 5))     # 1..4 and "5" for anything longer
```

## What to avoid

### Judging each repeat

`flag_tandem_repeats()` applies the thresholds: a run is flagged at 11 bp for a
homopolymer, 12 for a dinucleotide, 21 for a trinucleotide, and 50 for any unit. Those are
the values the HEOME Pf3D7 design used, and each is an argument -- `min_width_by_period`
is named by period, so `c("1" = 21, "2" = 21, "3" = 21)` holds homopolymers and
dinucleotides to the trinucleotide length, and `NULL` drops the per-period rules to flag on
total length alone.

```{r}
flagged <- flag_tandem_repeats(tr)
table(period = pmin(flagged$period, 4), avoid = flagged$avoid)
```

### Merging runs that flow into one another

Repeats flow into one another -- an `A` homopolymer into an `AT` run, one trinucleotide
into the next -- and the finders report each as its own record. `merge_tandem_repeats()`
joins records that overlap, abut, or lie within `gap` bases, and says what went into each
block:

```{r}
blocks <- merge_tandem_repeats(tr, gap = 10)
blocks[blocks$n_repeats >= 4, c("name", "width", "n_repeats", "repeat_units", "periods")][1:4, ]
```

The two finders often report one run twice under different spellings (`ATA` here, `TAT`
there). `dedupe = TRUE` drops a record lying inside another with the same unit up to
rotation or reverse complement; the blocks come out the same, and the unit list stops
naming both spellings.

### Judging the merged block

`tandem_repeats_to_avoid()` does merging and flagging together, and `rule` says which
width the thresholds see:

- `rule = "record"` (the default) judges each repeat on its own width, then keeps a merged
  block when any repeat in it was flagged, or when the block as a whole reaches 50 bp --
  a chain of individually harmless short repeats slips like one long one. This is the
  HEOME mask.
- `rule = "block"` merges first and judges each block on its **combined** width, held to
  the lowest threshold among the periods it contains. Any block with a homopolymer in it is
  held to 11 bp, whatever else is in it.

The difference is visible on two short repeats near each other. An 8 bp `A` run and a
9 bp `ATA` run 3 bp apart are each under their threshold, so the record rule keeps them;
merged within 10 bp they are one 20 bp stretch holding a homopolymer, and the block rule
avoids it:

```{r}
close <- data.frame(chrom = "Pf3D7_07_v3", start = c(100, 111), end = c(108, 120),
                    name = c("A_x8", "ATA_x3"))
tandem_repeats_to_avoid(close, gap = 10)                 # nothing
tandem_repeats_to_avoid(close, gap = 10, rule = "block")[, c("start", "end", "width", "periods")]
```

The masks these give for the genome:

```{r}
avoid <- tandem_repeats_to_avoid(tr)
avoid_block <- tandem_repeats_to_avoid(tr, gap = 10, rule = "block")
data.frame(rule = c("record", "block, gap 10"),
           blocks = c(nrow(avoid), nrow(avoid_block)),
           bases = c(sum(avoid$width), sum(avoid_block$width)))
```

The rest of this article uses `avoid`, the HEOME mask. Swap in `avoid_block`, or a stricter
`min_width_by_period`, and everything below runs the same.

## Workflow 1: genes, minus the repeats, to a BED

Take the genes you are after from `PF3D7_GENES`, subtract the mask with `bed_subtract()`,
and pad it so nothing you design ends right at a repeat's edge. What comes back is the
pieces of each gene that are clear of repeats, still carrying the gene's own columns:

```{r}
genes <- PF3D7_GENES[PF3D7_GENES$name %in% c("pfcrt", "pfdhfr", "pfkelch13"), ]
targets <- bed_subtract(genes, avoid, pad = 10)
targets[, c("name", "Pf3D7_chrom", "start", "end", "piece", "width")]
```

*pfcrt* comes back in many pieces, which is the point: it is a repeat-rich gene, and a
design that treats it as one interval will keep landing primers on the runs. `min_width`
drops pieces too short to be useful.

`write_bed6()` writes the six-column BED -- chromosome, start, end, name, score, strand --
using the reference's own chromosome names (`Pf3D7_chrom`, not the short `chrom` key) so
the file matches the FASTA. The score is the width unless the table has a `score` column,
and `meta_cols =` carries any other columns in a seventh `[field=value;]` column:

```{r}
bed <- file.path(tempdir(), "genes_no_repeats.bed")
write_bed6(targets, bed, meta_cols = c("gene_id", "piece"))
readLines(bed)[1:3]
```

Every piece of *pfcrt* is named `pfcrt` there. Tools that key on the name want it unique
within the file: `make_names_unique = TRUE` numbers repeats of a name in file order,
zero-padded to the count (`_01` to `_15`), and `name_is_coords = TRUE` names each interval
by its coordinates instead. The two combine, so identical coordinates written twice still
get distinct names:

```{r}
write_bed6(targets, bed, make_names_unique = TRUE)
readLines(bed)[2:3]
write_bed6(targets, bed, name_is_coords = TRUE)
readLines(bed)[2]
```

`write_bed()` is the three- or four-column form, for `bedtools` and `bcftools -R`.

## Workflow 2: genes, to exons, minus the repeats

Most of a gene's worst repeats sit in its introns, so a design will often want the exons
alone -- and the repeats that remain are in the exons, so the subtraction still follows.
`read_gff_features()` reads exons (or introns, CDS, or gene spans) out of a GFF3 as 0-based
intervals. `ensembl_gff_url()` points it at a whole released annotation:

```{r, eval = FALSE}
gff <- ensembl_gff_url("falciparum")         # Ensembl Protists, read straight from the web
exons <- read_gff_features(gff, "exon", ids = c("pfcrt", "pfdhfr", "pfkelch13"))
```

The package ships the annotation of six drug-resistance genes, verbatim from PlasmoDB,
which is what runs here; the coordinates are the same:

```{r}
gff <- system.file("extdata", "pf3d7_drug_gene_cds.gff", package = "plasgenomicsutilsR")
exons <- read_gff_features(gff, "exon", ids = c("pfcrt", "pfdhfr", "pfkelch13"))
exons[exons$name == "pfcrt", c("index", "start", "end", "width", "strand")]
```

`ids` takes gene ids, the GFF's own `Name` (`"CRT"`), the package's names (`"pfcrt"`, via
`genes = PF3D7_GENES`), or transcript ids. `index` numbers exons in transcript orientation,
so exon 1 of the minus-strand *pfkelch13* is the one with the highest coordinates;
`per = "transcript"` keeps isoforms apart instead of merging them.

Then the same subtraction, and the same BED:

```{r}
exon_targets <- bed_subtract(exons, avoid, pad = 10)
exon_targets[exon_targets$name == "pfcrt", c("index", "start", "end", "piece", "width")]

bed <- file.path(tempdir(), "exons_no_repeats.bed")
write_bed6(exon_targets, bed, meta_cols = c("gene_id", "index", "piece"))
readLines(bed)[1:3]
```

How much each route keeps:

```{r}
data.frame(route = c("gene minus repeats", "exons minus repeats"),
           pieces = c(nrow(targets), nrow(exon_targets)),
           bases = c(sum(targets$width), sum(exon_targets$width)))
```

`feature = "intron"` gives the gaps between exons if the introns are what you want to look
at, and `feature = "cds"` the coding parts only.
