---
title: "Avoiding tandem repeats"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Avoiding tandem repeats}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(plasgenomicsutilsR)
```

A polymerase slips on a short tandem repeat, and the shorter the repeating unit the shorter
a run it takes: a dozen `A`s is enough, a dozen bases of `AT` too, while a trinucleotide has
to run to about twenty before it matters and anything at all is trouble from fifty. Designing
targets -- amplicons, probes, windows to call in -- means knowing where those runs are and
keeping a few bases clear of them. This article goes from a repeat finder's output to a BED
of targets with the repeats cut out.

## The repeats

`pf3d7_tandem_repeats()` is every short tandem repeat in the Pf3D7 reference, found once by
Tandem Repeats Finder and an exhaustive search for perfect repeats of units up to 10 bp
(`?pf3d7_tandem_repeats` has the exact commands), and bundled so it never has to be run
again. `tandem_repeats()` reads the same kind of BED for any other genome.

```{r}
tr <- pf3d7_tandem_repeats()
nrow(tr)
tr[1:4, c("chrom", "start", "end", "width", "repeat_unit", "period", "copies")]
```

`period` is the length of the *true* repeating unit, whatever spelling the finder chose --
`AT`, `ATAT` and `ATATAT` all describe a dinucleotide run and all get `period = 2`. That is
the number the slippage rules key on:

```{r}
table(pmin(tr$period, 5))     # 1..4 and "5" for anything longer
```

## What to avoid

`flag_tandem_repeats()` applies the thresholds: a run is flagged at 11 bp for a
homopolymer, 12 for a dinucleotide, 21 for a trinucleotide, and 50 for any unit. Those are
the values used for the HEOME Pf3D7 design; each is an argument.

```{r}
flagged <- flag_tandem_repeats(tr)
table(period = pmin(flagged$period, 4), avoid = flagged$avoid)
```

Repeats flow into one another -- an `A` homopolymer into an `AT` run, one trinucleotide into
the next -- and the finders report each as its own record. `merge_tandem_repeats()` joins
records that overlap, abut, or lie within `gap` bases, and says what went into each block.
`tandem_repeats_to_avoid()` is the two steps together: a merged run is kept when any repeat
in it was flagged, or when the run as a whole reaches 50 bp, because a chain of individually
harmless short repeats slips like one long one.

```{r}
avoid <- tandem_repeats_to_avoid(tr)
nrow(avoid)
sum(avoid$width)          # bases masked
avoid[avoid$n_repeats > 3, c("name", "width", "n_repeats", "repeat_units", "periods")][1:3, ]
```

Stricter or looser is a matter of arguments: `min_width_by_period = c("1" = 9, "2" = 10)`
flags shorter runs, `min_width_by_period = NULL` flags on total length only, and
`gap = 5` also merges repeats separated by up to five bases.

## The targets

Take the genes you are after, subtract the mask with `bed_subtract()`, and pad it so
nothing you design ends right at a repeat's edge. What comes back is the pieces of each gene
that are clear of repeats, still carrying the gene's own columns:

```{r}
genes <- PF3D7_GENES[PF3D7_GENES$name %in% c("pfcrt", "pfdhfr", "pfkelch13"), ]
targets <- bed_subtract(genes, avoid, pad = 10)
targets[, c("name", "Pf3D7_chrom", "start", "end", "piece", "width")]
```

*pfcrt* comes back in many pieces, which is the point: it is a repeat-rich gene, and a
design that treats it as one interval will keep landing primers on the runs. `min_width`
drops pieces too short to be useful.

## Exons only

Most of a gene's worst repeats sit in its introns, so a design will often want the exons
alone. `read_gff_features()` reads them out of a GFF (the package ships the annotation of
six drug-resistance genes; `ensembl_gff_url()` points at a whole released one), and the flow
becomes gene, then exons, then the subtraction:

```{r}
gff <- system.file("extdata", "pf3d7_drug_gene_cds.gff", package = "plasgenomicsutilsR")
exons <- read_gff_features(gff, "exon", ids = c("pfcrt", "pfdhfr", "pfkelch13"))
exon_targets <- bed_subtract(exons, avoid, pad = 10)
exon_targets[exon_targets$name == "pfcrt", c("index", "start", "end", "piece", "width")]
```

`feature = "intron"` gives the gaps between exons, `"cds"` the coding parts only, and
`"gene"` each gene's whole span; `per = "transcript"` keeps isoforms apart instead of
merging them. Intervals come back 0-based half-open like everything else here, so they go
straight into `bed_subtract()` and `write_bed()`.

`write_bed()` writes the result, using the reference's own chromosome names
(`Pf3D7_chrom`, not the short `chrom` key) so the file matches the FASTA:

```{r, eval = FALSE}
write_bed(targets, "targets_no_repeats.bed")
```
